---
title: "Assessing Ecological Status"
author: "Anthony Sturbois/Miquel De Cáceres"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Assessing Ecological Status}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignettePackage{ecotraj}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction

### 1.1 Reference envelopes

### 1.2 About this vignette

In this vignette you will learn how to assess the ecological status of ecosystems by using a set of reference stations. 

First of all, we load `ecotraj`:

```{r load libraries, echo = T}
library(ecotraj)
```


## 2. Relationship with state reference envelope

## 2.1 GLOMEL example
Load GLOMEL data set
```{r}
data(glomel)
```

Subset compositional data matrix
```{r}
glomel_comp <- as.matrix(glomel[,!(names(glomel) %in% c("ID", "Ref", "Complementary"))])
rownames(glomel_comp) <- glomel$ID
```

Bray-Curtis distance
```{r}
glomel_bc <- vegan::vegdist(glomel_comp, method = "bray")
```

Get reference station IDs
```{r}
glomel_env <- glomel$ID[glomel$Ref]
```

```{r}
glomel_assess <- compareToStateEnvelope(glomel_bc, glomel_env, m=1.5)
```

Add interpretation
```{r}
glomel_assess$Interpretation <- glomel$Complementary
glomel_assess
```

Recalculate using confidence intervals
```{r}
glomel_assess <- compareToStateEnvelope(glomel_bc, glomel_env, m=1.5, nboot.ci = 1000)
glomel_assess
```

## 2.1 GLENAN example

```{r}
data(glenan)
```


Subset compositional data matrix
```{r}
glenan_comp <- as.matrix(glenan[,!(names(glenan) %in% c("Ref", "Group", "DistanceFromExtraction"))])
```

Bray-Curtis distance
```{r}
glenan_bc <- vegan::vegdist(glenan_comp, method = "bray")
```

Get reference station IDs
```{r}
glenan_env <- row.names(glenan)[glenan$Ref==1]
glenan_env
```

```{r}
glenan_assess <- compareToStateEnvelope(glenan_bc, glenan_env, m=1.5, nboot.ci = 1000)
glenan_assess$Interpretation <- glenan$Group
glenan_assess$DistanceFromExtraction <- glenan$DistanceFromExtraction
glenan_assess
```

```{r}
boxplot(glenan_assess$Q~  glenan_assess$Interpretation)
```

