---
title: "Introduction to Ecological Trajectory Analysis (ETA)"
author: "Miquel De Cáceres"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Introduction to Ecological Trajectory Analysis (ETA)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignettePackage{ecotraj}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction

### 1.1 What is Ecological Trajectory Analysis?

*Ecological Trajectory Analysis* (ETA) is a framework to analyze the dynamics of ecological entities (e.g. communities or ecosystems). The key aspect of ETA is that dynamics are represented using trajectories in a chosen multivariate space (noted using $\Omega$). These trajectories are then analyzed and compared geometrically. 

The ETA framework was presented for community ecology in De Cáceres et al (2019), and was extended with new metrics and visualisation modes in Sturbois et al. (2021a). Procedures of trajectory analysis can be applied to data beyond community data tables. For example, the same framework was applied to stable isotope data in Sturbois et al. (2021b). 

Since it can be applied to multiple target entities and multivariate spaces, we now refer to the framework as *Ecological Trajectory Analysis* and provide a package **ecotraj** that offers a set of functions to calculate metrics and produce plots. 

### 1.2 About this vignette

In this vignette you will learn how to conduct ETA using different package functions. In most of the vignette we describe how to study the trajectories of three target entities (i.e. sites, individuals, communities, etc.) that have been surveyed four times each. We use a small data set where trajectories occur in a space of two dimensions, so that geometric calculations can be followed more easily. In the last section a real example is presented.

First of all, we load `ecotraj`:

```{r load libraries, echo = T}
library(ecotraj)
```


## 2. Trajectory objects

### 2.1 Trajectory data

#### 2.1.1 Trajectory data items

To specify dynamics of a set of target entities, the following data items need to be distinguished: 

  a. A set of ecological states (i.e. coordinates in space $\Omega$) implicitly described using a **distance matrix** $\mathrm{D}$.
  b. A *character vector* specifying the **ecological entity** (i.e. sampling unit, community, ecosystem or individual) corresponding to each ecological state. Trajectory names are identified from unique values of entities.
  c. An *integer vector* specifying the **survey** (i.e. census) corresponding to the observation of each ecological state. This vector is important for survey order. If not provided, the order will be assumed to be incremental for each repetition of entity value.
  d. A *numeric vector* specifying the **survey time** corresponding the observation of each ecological state. This is needed for some metrics such as trajectory speed.

In ETA, target entities do not need to be surveyed synchronously nor the same number of times. 

#### 2.1.2 Example data set

Let us first define the vectors that describe the ecological entity and the survey of each observation:
```{r}
entities <- c("1","1","1","1","2","2","2","2","3","3","3","3")
surveys <- c(1,2,3,4,1,2,3,4,1,2,3,4)
```
We then define a matrix whose coordinates correspond to the set of ecological states observed. The number of rows in this matrix has to be equal to the length of vectors `entities` and `surveys`. For simplicity, we assume here that the ecological state space $\Omega$ has two dimensions:
```{r}
xy<-matrix(0, nrow=12, ncol=2)
xy[2,2]<-1
xy[3,2]<-2
xy[4,2]<-3
xy[5:6,2] <- xy[1:2,2]
xy[7,2]<-1.5
xy[8,2]<-2.0
xy[5:6,1] <- 0.25
xy[7,1]<-0.5
xy[8,1]<-1.0
xy[9:10,1] <- xy[5:6,1]+0.25
xy[11,1] <- 1.0
xy[12,1] <-1.5
xy[9:10,2] <- xy[5:6,2]
xy[11:12,2]<-c(1.25,1.0)
xy
```
The matrix of Euclidean distances $d$ between ecological states in $\Omega$ is then:
```{r} 
d <- dist(xy)
d
```

ETA is based on the analysis of information in the distance matrix $\mathrm{D} = [d]$. Therefore, it does not require explicit coordinates. This is an advantage because it allows the analysis to be conducted on arbitrary metric (or semi-metric) spaces. The choice of $d$ is left to the user and will depend on the problem at hand.

### 2.2 Defining trajectories

To perform ETA, we need to combine the distance matrix and the entity/survey information in a single object using function `defineTrajectories()`:

```{r}
x <- defineTrajectories(d, entities, surveys)
```

Note that `surveys` may be omitted, and in this case surveys for each entity are assumed to be ordered. The function returns an object (a list) of class `trajectories` that contains all the information for analysis:

```{r}
class(x)
```

This object contains two elements:
```{r}
names(x)
```
Element `d` contains the input distance matrix, whereas `metadata` is a data frame including information of observations:

```{r}
x$metadata
```

Column `sites` identifies the ecological entities (calling them `sites` is an inherited notation from the original framework of community trajectory analysis). Note that columns `surveys` and `times` have exactly the same values. This happens because we did not supplied a vector for `times` so that surveys are assumed to happen every time step (of whatever units). Moreover, the `surveys` vector itself can be omitted in calls to `defineTrajectories()`. If so, the function will (correctly, in this case) interpret that every repetition of a given entity corresponds to a new survey:

```{r}
x <- defineTrajectories(d, entities)
x$metadata
```

Let us assume the following sampling times, in units of years:
```{r}
times <- c(1.0,2.2,3.1,4.2,1.0,1.5,2.8,3.9,1.6,2.8,3.9,4.3)
```

The call to `defineTrajectories()` using all the information would be:
```{r}
xt <- defineTrajectories(d, entities, surveys, times)
xt$metadata
```

Note that in `x` all entities have been surveyed in the exact same times. The resulting trajectories are called **synchronous**. In contrast, in `xt` the entities have been surveyed at different times, so that trajectories are **non-synchronous**. In the following, we will use `xt` whenever this distinction is relevant.

### 2.3 Subsetting trajectories
At some point in the ETA, one may desire to focus on particular trajectories or surveys. Function `subsetTrajectory()` allows subsetting objects of class `trajectories`, For example, we can decide to work with the trajectories of the second and third entities (sites):
```{r}
x23 <- subsetTrajectories(x, site_selection = c("2", "3"))
x23
```

We can decide to focus on the last three surveys:
```{r}
x23s <- subsetTrajectories(x, 
                           site_selection = c("2", "3"), 
                           survey_selection = c(2, 3, 4))
x23s
```
You will notice that `surveys` have been renumbered (but original `times` are not modified). This illustrates that the vector `surveys` is only used to indicate the survey order within each trajectory.


### 2.4 Displaying trajectories

To begin our analysis of the three trajectories, we display them in an ordination space, using function `trajectoryPCoA()`. Since $\Omega$ has only two dimensions in this example, the Principal Coordinates Analysis (PCoA) on $d$ displays the complete space:

```{r pcoa, fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
trajectoryPCoA(x, traj.colors = c("black","red", "blue"), lwd = 2,
               survey.labels = T)
legend("topright", col=c("black","red", "blue"), 
       legend=c("Entity 1", "Entity 2", "Entity 3"), bty="n", lty=1, lwd = 2)
```

While trajectory of entity '1' (black arrows) is made of three segments of the same length and direction, trajectory of entity '2' (red arrows) has a second and third segments that bend and are shorter than that of the second segment of entity '1'. Trajectory of entity '3' includes a stronger change in direction and shorter segments. 

As this example has two dimensions and we used Euclidean distance, the same plot (albeit rotated) can be straightforwardly obtained using matrix `xy` and function `trajectoryPlot()`:

```{r xy, fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
trajectoryPlot(xy, entities, surveys, traj.colors = c("black","red", "blue"), lwd = 2,
               survey.labels = T)
legend("topright", col=c("black","red", "blue"), 
       legend=c("Entity 1", "Entity 2", "Entity 3"), bty="n", lty=1, lwd = 2)
```

While `trajectoryPCoA()` uses PCoA (also known as classical Multidimensional Scaling) to display trajectories, users can display ecosystem trajectories using other ordination techniques such as metric Multidimensional Scaling (mMDS; see function *mds* of package **smacof**) or non-metric MDS (nMDS; see function *metaMDS* in package **vegan** or function *isoMDS* in package **MASS**). Function `trajectoryPlot()` will help drawing arrows between segments to represent trajectories on the ordination space given by any of these methods. 

Functions `trajectoryPCoA()`and `trajectoryPlot()` can be used to display a subset of trajectories if we combine them with function `subsetTrajectories()`:

```{r, fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
trajectoryPCoA(subsetTrajectories(x, site_selection = c("2", "3")), 
               traj.colors = c("red", "blue"), lwd = 2,
               survey.labels = T)
legend("topright", col=c("red", "blue"), 
       legend=c("Entity 2", "Entity 3"), bty="n", lty=1, lwd = 2)
```


## 3. Trajectory metrics

One may be interested in studying the geometric properties of particular trajectories. This is illustrated in this section

### 3.1 Changes in ecological state

Several metrics are related to the magnitude of temporal changes in state. For example, one can obtain the length of trajectory segments and the total path length using function `trajectoryLengths()`:
```{r} 
trajectoryLengths(x)
```

Since the difference between `x` and `xt` is in observation times, we will obtain the same result when calculating lengths on `xt`:

```{r} 
trajectoryLengths(xt)
```

When observation times are available, it may be of interest to calculate segment or trajectory speeds. One can obtain the speed of trajectory segments and the total path speed using function `trajectorySpeeds()`:

```{r} 
trajectorySpeeds(x)
```
Note that the units of lengths and speeds will depend on the definition of the $\Omega$ space and, in the latter case, on the units of `times`. Trajectory speeds are affected by observation times and, since in `x` times are consecutive integers, segment speeds are equal to segment lengths (but not the average trajectory speed). We will obtain a different result for speeds with `xt`:

```{r} 
trajectorySpeeds(xt)
```

Finally, one may calculate the internal variation of states within each trajectory using `trajectoryInternalVariation()`:
```{r}
trajectoryInternalVariation(x)
```
The function returns the (absolute or relative) contribution of each observation to the internal variation, the total sum of squares and an unbiased estimation of internal variance. Note that in this example the third (more curved) trajectory has lower internal variation, compared to the first and second (straighter) ones.

### 3.2 Changes in direction

In CTA, angles are measured using triplets of time-ordered ecological states (a pair of consecutive segments is an example of such triplets). As matrix $\mathrm{D}$ may represent a space $\Omega$ of multiple dimensions, angles cannot be calculated with respect to a single plane. Instead, each angle is measured on the plane defined by each triplet. Zero angles indicate that the three points (e.g. the two consecutive segments) are in a straight line. The larger the angle value, the more is trajectory changing in direction. Mean and standard deviation statistics of angles are calculated according to circular statistics. Function `trajectoryAngles()` allows calculating the angles between consecutive segments:

```{r} 
trajectoryAngles(x)
```

While entity '1' follows a straight path, angles are > 0 for trajectories of entity '2' and '3', denoting the change in direction. In this case, the same information could be obtained by inspecting the previous plots, but in a case where $\Omega$ has many dimensions, the representation will correspond to a reduced (ordination) space and hence, angles and lengths in the plot will not correspond exactly to those of functions `trajectoryLengths()` and `trajectoryAngles()`, which take into account the complete $\Omega$ space. 

Angles can be calculated not only for all consecutive segments but for all four triplets of ordered ecological states, whether of consecutive segments or not (i.e., between points 1-2-3, 1-2-4, 1-3-4 and 2-3-4). This is done by specifying `all=TRUE` in `trajectoryAngles()`:

```{r} 
trajectoryAngles(x, all=TRUE)
```

The *mean resultant length* of circular statistics (column `rho` of the previous result), which takes values between 0 and 1, can be used to assess the degree of homogeneity of angle values and it will take a value of 1 if all angles are the same. This approach, however, uses only angular information and does not take into account the length of segments. 

To measure the overall directionality of an ecosystem trajectory (i.e. if the path consistently follows the same direction in $\Omega$ ), we recommend using another statistic that is sensitive to both angles and segment lengths and is implemented in function `trajectoryDirectionality()`:
```{r} 
trajectoryDirectionality(x)
```  

As known from previous plots, trajectory of entity '2' is less straight than trajectory of entity '1' and that of entity '3' has the lowest directionality value. By default the function only computes a descriptive statistic, i.e. it does not perform any statistical test on directionality. A permutational test can be performed, but this feature is experimental and needs to be tested before recommendation.

### 3.2 Assessing multiple metrics at once
It is possible to assess multiple trajectory metrics in one function call to `trajectoryMetrics()`. This will only provide metrics that apply to the whole trajectory path: 
```{r}
trajectoryMetrics(x)
```

If we calculate metrics on `xt` we will confirm that only trajectory speeds are affected by observation times:
```{r}
trajectoryMetrics(xt)
```

Another function, called `trajectoryWindowMetrics()` calculates trajectory metrics on moving windows over trajectories, but will not be illustrated here.

## 4. Comparing trajectories

### 4.1 Relative positions within trajectories
Ecological states occupy a position within their trajectory that depends on the total path length of the trajectory (see Fig. 2 of De Cáceres et al. 2019). By adding the length of segments prior to a given state and dividing the sum by the total length of the trajectory we obtain the relative position of the ecological state. Function `trajectoryProjection()` allows obtaining the relative position of each ecological state of a trajectory. To use it for this purpose one should use as parameters the distance matrix between states and the indices that conform the trajectory, which have to be entered twice. For example for the two example trajectories we would have:

```{r} 
trajectoryProjection(d, 1:4, 1:4)
```  

If we inspect the relative positions of the points in the trajectory of entity '2', we find than the second and third segments have relative positions larger than 1/3 and 2/3, respectively, because the second and third segments are shorter:
```{r} 
trajectoryProjection(d, 5:8, 5:8)
```

Function `trajectoryProjection()` can also be used to perform an **orthogonal projection** of arbitrary ecological states onto a given reference trajectory. For example we can study the projection of third state of the trajectory of entity '2' (i.e. state 7) onto the trajectory of entity '1' (i.e. states 1 to 4), which happens to be in the half of the trajectory:

```{r} 
trajectoryProjection(d, 7, 1:4)
```  
If we project the points of the trajectory of entity '3' onto the trajectory of entity '1' we see how the curved path of entity '3' projects its fourth point to the same relative position as its second point.

```{r} 
trajectoryProjection(d, 9:12, 1:4)
```  

### 4.2 Trajectory shifts

Sometimes different ecosystems follow the same or similar trajectory but with different speeds, or with an observations starting at a different point in the dynamic sequence. We can quantify those differences using function `trajectoryShifts()`, which internally uses orthogonal projection. To illustrate this function, we will first build a small data set of three linear trajectories, but where the second and the third are modified:

```{r}
#Description of entities and times
entities3 <- c("1","1","1","1","2","2","2","2","3","3","3","3")
times3 <- c(1,2,3,4,1,2,3,4,1,2,3,4)
  
#Raw data table
xy3<-matrix(0, nrow=12, ncol=2)
xy3[2,2]<-1
xy3[3,2]<-2
xy3[4,2]<-3
xy3[5:8,1] <- 0.25
xy3[5:8,2] <- xy3[1:4,2] + 0.5 # States are all shifted with respect to entity "1"
xy3[9:12,1] <- 0.5
xy3[9:12,2] <- xy3[1:4,2]*1.25  # 1.25 times faster than entity "1"
```

We can see the differences graphically:

```{r, fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
trajectoryPlot(xy3, entities3,  
               traj.colors = c("black","red", "blue"), lwd = 2)
legend("topright", col=c("black","red", "blue"), 
       legend=c("Trajectory 1", "Trajectory 2", "Trajectory 3"), bty="n", lty=1, lwd = 2)
```


We now build the usual `trajectories` object:
```{r}
x3 <- defineTrajectories(dist(xy3), entities3, times = times3)
```

We can check that indeed the third trajectory is faster using:

```{r}
trajectorySpeeds(x3)
```
Function `trajectoryShifts()` allows comparing different observations to a reference trajectory. For example we can compare trajectory for entities "1" and "2":

```{r}
trajectoryShifts(subsetTrajectories(x3, c("1","2")))
```
Where we see that the observations of trajectory "2" correspond to states of trajectory "1" at 0.5 time units later in time. Surveys with missing values indicate that the projection of the target state cannot be determined (because the reference trajectory is too short).

We can also compare trajectories "1" and "3":
```{r}
trajectoryShifts(subsetTrajectories(x3, c("1","3")))
```

Here we see that shifts increase progressively, indicating the faster speed of trajectory "3".

### 4.3 Trajectory convergence/divergence

When trajectories are synchronous, one can study their symmetric convergence or divergence (see Fig. 3a of De Cáceres et al. 2019). Function `trajectoryConvergence()` allows performing tests of convergence based on the trend analysis of the sequences of distances between points of the two trajectories (i.e. first-first, second-second, ...):

```{r} 
trajectoryConvergence(x, type = "pairwise.symmetric")
```  

The function performs the Mann-Whitney trend test. Values of the statistic ('tau') larger than 0 correspond to trajectories that are diverging, whereas values lower than 0 correspond to trajectories that are converging. By setting `type = "pairwise.asymmetric"` the convergence test becomes asymmetric (see Figs. 3b and 3c of De Cáceres et al. 2019). In this case the sequence of distances *between every point of one trajectory and the other*:

```{r} 
trajectoryConvergence(x, type = "pairwise.asymmetric")
```  

The asymmetric test is useful to determine if one trajectory is becoming closer to the other or if it is departing from the other. The asymmetric test can be applied on non-synchronous trajectories. Finally, if the trajectories have been surveyed synchronously, one can also perform a global test of convergence/divergence between trajectories, using `type = "multiple"`:

```{r} 
trajectoryConvergence(x, type = "multiple")
```  

In this case we are testing whether the average distance between ecological states corresponding to the same observation time is increasing or decreasing with time. In all these tests trajectories are diverging (as indicated by the positive tau values) but the tests are not statistically significance due to the small number of surveys.

### 4.4 Distances between segments and between trajectories

The ETA framework allows quantifying the resemblance in the dynamics of target entities by assessing the dissimilarity of their corresponding trajectories. Broadly speaking, dissimilarity between trajectories will be influenced both by differences in ecological states that are constant in time and differences that arise from temporal changes. To focus on the second, distances between trajectories can be calculated after centering them (i.e. after bringing all trajectories to the center of the $\Omega$ space). This is done using function `centerTrajectories()`, which returns a new dissimilarity matrix and is illustrated in article "Transforming trajectories". 

#### 4.4.1 Distances between segments

For some trajectory dissimilarity coefficients, one intermediate step is the calculation of distances between directed segments (see Fig. 4 of De Cáceres et al. 2019), which can be obtained by calling function `segmentDistances`:
```{r} 
Ds <- segmentDistances(x)$Dseg
Ds
```  

Distances between segments are affected by differences in both *position*, *size* and *direction*. Hence, among the six segments of this example, the distance is maximum between the last segment of trajectory '1' (named `1[3-4]`) and the first segment of trajectory '3' (named `3[1-2]`). 

One can display distances between segments in two dimensions using mMDS.
```{r, echo=T, fig=TRUE, fig.height=5, fig.width=5, fig.align = "center"}
mMDS <- smacof::mds(Ds)
mMDS
xret <- mMDS$conf
plot(xret, xlab="axis 1", ylab = "axis 2", asp=1, pch=21,
     bg=c(rep("black",3), rep("red",3), rep("blue",3)), 
     xlim=c(-1.5,1), ylim=c(-1,1.5))
text(xret, labels=rep(paste0("s",1:3),3), pos=1)
legend("topleft", pt.bg=c("black","red","blue"), pch=21, bty="n", legend=c("Trajectory 1", "Trajectory 2", "Trajectory 3"))
```

#### 4.4.2 Distances between trajectories

Distances between segments are internally calculated when comparing whole trajectories using function `trajectoryDistances()`. Here we show the dissimilarity between the two trajectories as assessed using either the *Hausdorff* distance (equal to the maximum distance between directed segments; see eq. 8 in De Cáceres et al. 2019), the *segment path distance* (Besse et al, 2016), the *directed segment path distance* (see eq. 13 in De Cáceres et al. 2019) or the *time-sensitive path distance* (unpublished): 
```{r} 
trajectoryDistances(x, distance.type = "Hausdorff")
trajectoryDistances(x, distance.type = "SPD")
trajectoryDistances(x, distance.type = "DSPD")
trajectoryDistances(x, distance.type = "TSPD")
```  

SPD, DSPD and TSPD are symmetrized by default. To calculate non-symmetric distances one uses, for example (see eq. 11 in De Cáceres et al. 2019):
```{r} 
trajectoryDistances(x, distance.type = "DSPD", symmetrization = NULL)
```  

A detailed comparison of trajectory dissimilarity indices can be found in article "Distance metrics for trajectory resemblance".

When estimating dissimilarities between a set of trajectories one is building a new space (noted as $\Omega_T$). One can display the location of trajectories in two dimensions using mMDS.
```{r, echo=T, fig=TRUE, fig.height=5, fig.width=5, fig.align = "center"}
mMDS <- smacof::mds(trajectoryDistances(x, distance.type = "TSPD"))
mMDS
xret <- mMDS$conf
plot(xret, xlab="axis 1", ylab = "axis 2", asp=1, pch=21,
     bg=c("black", "red", "blue"), 
     xlim=c(-1.0,1), ylim=c(-1,1.0))
legend("topleft", pt.bg=c("black","red","blue"), pch=21, bty="n", legend=c("Trajectory 1", "Trajectory 2", "Trajectory 3"))
```


### 4.5 Dynamic variation

One may be interested in knowing how much diverse are a set of trajectories, and which entities follow dynamics more distinct from others. We refer to the diversity of trajectories as **dynamic variation**., and these questions can be addressed using function `dynamicVariation()`, for example:

```{r} 
dynamicVariation(x)
```  

Analogously to `trajectoryInternalVariation()`, function `dynamicVariation()` returns the sum of squares of dynamic variation, an unbiased dynamic variance estimator and the relative contribution of individual trajectories to the overall sum of squares. Function `dynamicVariation()`, makes internal calls to `trajectoryDistances()`, which means that we may get slightly different results if we change the trajectory dissimilarity coefficient:

```{r} 
dynamicVariation(x, distance.type = "TSPD")
```  

## 5. Real example: structural dynamics in permanent plots

In this example we analyze the dynamics of 8 permanent forest plots located on slopes of a valley in the New Zealand Alps. The study area is mountainous and centered on the Craigieburn Range (Southern Alps), South Island, New Zealand (see map in Fig. 8 of De Cáceres et al. 2019).  Forests plots are almost monospecific, being the mountain beech (*Fuscospora cliffortioides*) the main dominant tree species. Previously forests consisted of largely mature stands, but some of them were affected by different disturbances during the sampling period (1972-2009) which includes 9 surveys. We begin our example by loading the data set, which includes 72 plot observations:
```{r load avoca, echo=T}
data("avoca")
```

Community data is in form of an object `stratifiedvegdata`. To account for differences in tree diameter, while emphasizing regeneration, the data contains individual counts to represent tree abundance and trees are classified in 19 quadratic diameter bins (in cm): {(2.25, 4], (4, 6.25], (6.25, 9], ... (110.25, 121]}. The data set also includes vectors `avoca_surveys` and `avoca_sites` that indicate the survey and forest plot corresponding to each forest state.

Before starting ETA, we have to use function `vegdiststruct` from package **vegclust** to calculate distances between forest plot states in terms of structure and composition (see De Cáceres M, Legendre P, He F (2013) Dissimilarity measurements and the size structure of ecological communities. Methods Ecol Evol 4:1167–1177. https://doi.org/10.1111/2041-210X.12116):
```{r distance, echo=TRUE}
avoca_D_man <- vegclust::vegdiststruct(avoca_strat, 
                                       method="manhattan", 
                                       transform = function(x){log(x+1)})
```

Distances in `avoca_D_man` are calculated using the Manhattan metric, after applying a logarithm transformation to abundance data. 

### 5.1 Defining trajectories

We start by defining our trajectories, which implies combining the information about distances, entities (sites) and observation times. The nine surveys were conducted in years 1970/72, 1974, 1978, 1983, 1987, 1993, 1999, 2004 and 2009. Since plots have been surveyed rather synchronously, we can define observation times using:

```{r}
years <- c(1971, 1974, 1978, 1983, 1987, 1993, 1999, 2004, 2009)
avoca_times <- years[avoca_surveys]
```

And then define trajectories using:

```{r}
avoca_x <- defineTrajectories(d = avoca_D_man,  
                              sites = avoca_sites, 
                              times = avoca_times)
```

In this dataset trajectories , as indicated by function `is.synchronous()`.
```{r}
is.synchronous(avoca_x)
```

On the other hand, the Manhattan metric produces a distance matrix (and $\Omega$ space) that is indeed metric:
```{r}
is.metric(avoca_x)
```


### 5.2 Displaying trajectories in PCoA

The distance matrix `avoca_D_man` conforms our definition of $\Omega$. We use `trajectoryPCoA()` to display the relations between forest plot states in this space and to draw the trajectory of each plot:

```{r avoca_pcoa, fig=TRUE, fig.height=5, fig.width=5, fig.align = "center"}
trajectoryPCoA(avoca_x,
               traj.colors = RColorBrewer::brewer.pal(8,"Accent"), 
               axes=c(1,2), length=0.1, lwd=2)
legend("topright", bty="n", legend = 1:8, col = RColorBrewer::brewer.pal(8,"Accent"), lwd=2)
```

Note that in this case, the full $\Omega$ includes more than two dimensions, and PCoA is representing 43% of total variance (correction for negative eigenvalues is included in the call to `cmdscale` from `trajectoryPCoA()`), so one has to be careful when interpreting trajectories visually.

Another option is to use mMDS to represent trajectories, which in this case produces a similar result:

```{r avoca_mmds, fig=TRUE, fig.height=5, fig.width=5, fig.align = "center"}
mMDS <- smacof::mds(avoca_D_man)
mMDS
trajectoryPlot(mMDS$conf,  avoca_sites, avoca_surveys,
               traj.colors = RColorBrewer::brewer.pal(8,"Accent"), 
               axes=c(1,2), length=0.1, lwd=2)
legend("topright", bty="n", legend = 1:8, col = RColorBrewer::brewer.pal(8,"Accent"), lwd=2)
```

```{r int1, echo=FALSE}
plotTrajDiamDist<-function(cli = 7) {
l = colnames(avoca_strat[[1]])
ncl = 14
m197072= avoca_strat[avoca_surveys==1][[cli]]["NOTCLI",2:ncl]
m197072[m197072<1] = NA
m1974 = avoca_strat[avoca_surveys==2][[cli]]["NOTCLI",2:ncl]
m1974[m1974<1] = NA
m1978 = avoca_strat[avoca_surveys==3][[cli]]["NOTCLI",2:ncl]
m1978[m1978<1] = NA
m1983 = avoca_strat[avoca_surveys==4][[cli]]["NOTCLI",2:ncl]
m1983[m1983<1] = NA
m1987 = avoca_strat[avoca_surveys==5][[cli]]["NOTCLI",2:ncl]
m1987[m1987<1] = NA
m1993 = avoca_strat[avoca_surveys==6][[cli]]["NOTCLI",2:ncl]
m1993[m1993<1] = NA
m1999 = avoca_strat[avoca_surveys==7][[cli]]["NOTCLI",2:ncl]
m1999[m1999<1] = NA
m2004 = avoca_strat[avoca_surveys==8][[cli]]["NOTCLI",2:ncl]
m2004[m2004<1] = NA
m2009 = avoca_strat[avoca_surveys==9][[cli]]["NOTCLI",2:ncl]
m2009[m2009<1] = NA


plot(m197072, type="l", ylim=c(1,200), log="y",
       xlab="", ylab="Number of individuals (log)", main=paste0("Trajectory ",cli), 
       axes=FALSE, col=gray(0.8), lwd=2)
axis(2, las=2)
axis(1, at=1:(ncl-1), labels=l[2:ncl], las=2)
lines(m1974, col=gray(0.7), lwd=2)
lines(m1978, col=gray(0.6), lwd=2)
lines(m1983, col=gray(0.5), lwd=2)
lines(m1987, col=gray(0.4), lwd=2)
lines(m1993, col=gray(0.3), lwd=2)
lines(m1999, col=gray(0.2), lwd=2)
lines(m2004, col=gray(0.1), lwd=2)
lines(m2009, col=gray(0), lwd=2)
legend("topright", bty="n", lwd=2,col=gray(seq(0.8,0, by=-0.1)), legend=c("1970/72","1974","1978","1983", "1987", "1993","1999","2004","2009"))
}
```

One can inspect specific trajectories using `subsetTrajectories()`. This allows getting a better view of particular trajectories, here that of forest plot '3':
```{r trajectory_3, fig.height=4, fig.width=9, fig.align = "center"}
oldpar <- par(mfrow=c(1,2))
trajectoryPCoA(subsetTrajectories(avoca_x, "3"),
               length=0.1, lwd=2, time.labels = TRUE)
plotTrajDiamDist(3)
par(oldpar)
```

In the right hand, we added a representation of the change in the mountain beech tree diameter distribution through time for trajectory of forest plot '3'. The dynamics of this plot include mostly growth, which results in individuals moving from one diameter class to the other. The whole trajectory looks mostly directional. Let's now inspect the trajectory of forest plot '4':
```{r trajectory_4, echo=T, fig.height=5, fig.width=9, fig.align = "center"}
oldpar <- par(mfrow=c(1,2))
trajectoryPCoA(subsetTrajectories(avoca_x, "4"),
               length=0.1, lwd=2, time.labels = TRUE)
plotTrajDiamDist(4)
par(oldpar)
```

This second trajectory is less straight and seems to include a turn by the end of the sampling period, corresponding to the recruitment of new saplings.



### 5.3 Trajectory lengths, angles and overall directionality

While trajectory lengths and angles can be inspected visually in ordination diagrams, it is better to calculate them using the full $\Omega$ space (i.e., from matrix `avoca_D_man`). Using function `trajectoryLengths()` we can see that the trajectory of forest plot '4' is lengthier than that of plot '3', mostly because includes a lengthier last segment (i.e. the recruitment of new individuals):

```{r trajectory lengths}
trajectoryLengths(avoca_x)
```

If we calculate the angles between consecutive segments (using function `trajectoryLengths`) we see that indeed the trajectory of '3' is rather directional, but the angles of trajectory of '4' are larger, on aveerage:

```{r trajectory angles, echo=T}
avoca_ang <- trajectoryAngles(avoca_x)
avoca_ang
```

By calling function `trajectoryDirectionality()` we can confirm that the trajectory for forest plot '4' is less straight than that of plot '3':

```{r trajectory directionality, echo=T}
avoca_dir <- trajectoryDirectionality(avoca_x)
avoca_dir
```

The following code displays the relationship between the statistic in `trajectoryDirectionality()` and the mean resultant vector length that uses angular information only and assesses the constancy of angle values:

```{r, echo=TRUE, fig = TRUE, fig.height=5, fig.width=5, fig.align="center"}
avoca_rho <- trajectoryAngles(avoca_x, all=TRUE)$rho
plot(avoca_rho, avoca_dir, xlab = "rho(T)", ylab = "dir(T)", type="n")
text(avoca_rho, avoca_dir, as.character(1:8))
```

### 5.4 Convergence between trajectories

We may ask if structure in forest plots is becoming more similar with time. This question can be addressed using an overall test of convergence, which we can do because trajectories are synchronous:

```{r}
trajectoryConvergence(avoca_x, type="multiple")
```

In this case we obtain that tau is decreasing and the test is significant, which indicates that forest structures are overall converging. This general trend may not be true for specific pairs of plots. Converge/divergence between pairs of plots would be assessed using:

```{r}
trajectoryConvergence(avoca_x, type="pairwise.symmetric")
```

Inspecting these results we can conclude that several pairs of plots are significantly converging (notably with plot '8'), but there is also a significant divergence between plots '4' and '5'. We can display this divergence graphically using:

```{r trajectories_4_5_plot, fig.height=5, fig.width=5, fig.align = "center"}
trajectoryPCoA(subsetTrajectories(avoca_x, c("4", "5")),
               traj.colors = RColorBrewer::brewer.pal(8,"Accent")[4:5], 
               axes=c(1,2), length=0.1, lwd=2)
legend("topright", bty="n", legend = 4:5, col = RColorBrewer::brewer.pal(8,"Accent")[4:5], lwd=2)
```

To interpret this result we can compare the corresponding changes in diameter distribution:

```{r trajectories_4_5_diam, echo=T, fig.height=5, fig.width=9, fig.align = "center"}
oldpar <- par(mfrow=c(1,2))
plotTrajDiamDist(4)
plotTrajDiamDist(5)
par(oldpar)
```

Apparently, the divergence would be explained by the fact that while plot '4' evolves towards a more regular structure of medium/large trees  (i.e. from more diverse tree size distribution towards less diverse one), plot '5' maintains an irregular structure (i.e. diverse tree size distribution) throughout the years thanks to a greater sapling ingrowth.

### 5.5 Distances between trajectories

We can calculate the resemblance between forest plot trajectories using `trajectoryDistances()`:
```{r}
avoca_D_traj_man <- trajectoryDistances(avoca_x, distance.type="DSPD")
print(round(avoca_D_traj_man,3))
```
The closest trajectories are those of plots '1' and '2'. They looked rather close in *position* in the PCoA ordination of $\Omega$ with all trajectories, so probably it is position, rather than *shape* which has influenced this low value. The next pair of similar trajectories are those of the '3'-'5' pair. We can again use mMDS to produce an ordination of resemblances between trajectories:

```{r avoca_DT_PCoA, echo=TRUE, fig = TRUE, fig.height=5, fig.width=5, fig.align="center"}
mMDS<-smacof::mds(avoca_D_traj_man)
mMDS
x<-mMDS$conf[,1]
y<-mMDS$conf[,2]
plot(x,y, type="p", asp=1, xlab=paste0("Axis 1"), 
     ylab=paste0("Axis 2"), col="black",
     bg= RColorBrewer::brewer.pal(8,"Accent"), pch=21)
text(x,y, labels=1:8, pos=1)
```



### 5.6 Dynamic variation

To determine which forest plots have more unique structural dynamics, we can use function `dynamicVariation()`:

```{r} 
dynamicVariation(avoca_x, distance.type="DSPD")
```  

We see that forest plots '3', '4', '5' and '7' contribute much less to overall variation in dynamics than the others. These plots are the same that were located closer to the center of the previous ordination plot. The more unique plot is '8', which we can isolate and inspect using:

```{r trajectory_8, echo=T, fig.height=5, fig.width=9, fig.align = "center"}
oldpar <- par(mfrow=c(1,2))
trajectoryPCoA(subsetTrajectories(avoca_x, "8"),
               length=0.1, lwd=2, time.labels = TRUE)
plotTrajDiamDist(8)
par(oldpar)
```

Apparently, the distinctiveness of plot '8' from the remaining stems from its very low number of trees at the beginning and the large amount of regeneration. This structural dynamics would be rather different from that of other plots that have more adults in the beginning and less amount of regeneration.

## 6. References

Besse, P., Guillouet, B., Loubes, J.-M. & François, R. (2016). Review and perspective for distance based trajectory clustering. IEEE Trans. Intell. Transp. Syst., 17, 3306–3317.

De Cáceres M, Coll L, Legendre P, Allen RB, Wiser SK, Fortin MJ, Condit R & Hubbell S. (2019). Trajectory analysis in community ecology. Ecological Monographs 89, e01350.

Sturbois, A., De Cáceres, M., Sánchez-Pinillos, M., Schaal, G., Gauthier, O., Le Mao, P., Ponsero, A., & Desroy, N. (2021a). Extending community trajectory analysis : New metrics and representation. Ecological Modelling 440: 109400. https://doi.org/10.1016/j.ecolmodel.2020.109400.

Sturbois, A., Cucherousset, J., De Cáceres, M., Desroy, N., Riera, P., Carpentier, A., Quillien, N., Grall, J., Espinasse, B., Cherel, Y., Schaal, G. (2021b). Stable Isotope Trajectory Analysis (SITA) : A new approach to quantify and visualize dynamics in stable isotope studies. Ecological Monographs, 92, e1501. https://doi.org/10.1002/ecm.1501. 
