---
title: "Introduction to Cyclical Ecological Trajectory Analysis (CETA)"
author: "Nicolas Djeghri/Miquel De Cáceres"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Introduction to Cyclical Ecological Trajectory Analysis (CETA)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignettePackage{ecotraj}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction

### 1.1 Going round in cycles, but go somewhere! What CETA does, and where it sits
*Cyclical Ecological Trajectory Analysis* (CETA) is an extension of *Ecological Trajectory Analysis* (ETA) allowing it to handle trajectories presenting regular cyclical dynamics (e.g. seasons, day-night cycle). 
Before starting, it is important to clarify what CETA does and what it does not, as well as what we mean by *regular cyclical dynamics*.  

Unlike many statistical procedures dedicated to time series, CETA does not aims to detect cyclicity. Rather, CETA aims at describing it and gaining insights into long-term changes in cyclical dynamics. Cyclicity must be known *a priori* from knowledge of the system under study or from another statistical analysis (e.g. eigenvector maps).  

By "regular cyclical dynamics" we mean cycles that are ‒ *most often* ‒ the product of the very regularly paced astronomic forcings (seasons, day-night cycles, tides). Describing seasonal dynamics was in fact the prime motivation in designing CETA. CETA is therefore not designed to address many other dynamics that ecologists would refer to as "cyclical" such as disturbance-recovery cycles, because the time duration of those cycles is not fixed (i.e. they are not periodic). These may be better addressed with clever use of the more general ETA framework. The primary research field we envision for CETA lies in a middle ground between phenology, community, and ecosystem ecology.

### 1.2 About this vignette
In this vignette you will learn how to use the specific functions designed for CETA, and how to combine them with the wider ETA framework to obtain metrics describing cyclical trajectories.  
Let's first call the package:
```{r}
library(devtools)
load_all("C:/Users/djeghri/Documents/GitHub/ecotraj")
#library(ecotraj)
```

## 2. General approach of CETA

### 2.1 The vocabulary of CETA
CETA is perhaps a bit heavy-handed on the vocabulary. We have two words for time and recognize no less than three different types of trajectories! Let's start with the way we refer to time: CETA distinguishes "**times**", representing linear time, and "**dates**" representing circular time. For instance, the 17th of December recurs every year, in that sense, we call it a date. However, the 17th of December 2009 occurred only once ([and BBC radio is grateful](https://www.youtube.com/watch?v=AmPN8MPw2dk)). Then, there is the three types of trajectories that CETA distinguishes and characterizes:

* **Cyclical trajectories**: Long trajectories presenting cyclical dynamics of a given periodicity (e.g. annual). They are the basis of CETA. They also constitute the inputs from which the two other types of trajectories will be derived.
* **Cycles**: Sub-divisions of a cyclical trajectory of duration equal to its periodicity. 
* **Fixed-date trajectories**: Trajectories joining the ecological states sampled at the same date in a given cyclical trajectory (for instance, in a multi-annual monthly sampled time series, the trajectory joining the ecological states obtained for March of year 1, March of year 2, etc...).

All this vocabulary helps CETA being a little more time-explicit than regular ETA. Let's build a toy dataset with **one cyclical trajectory** composed of **three cycles** to make this more visual:
```{r}
#Let's define our toy sampling times:
timesToy <- 0:30 #The sampling times of the time series
cycleDurationToy <- 10 #The duration of the cycles (i.e. the periodicity of the time series)
datesToy <- timesToy%%cycleDurationToy #The dates associated to each times

#And state where the sampling occurred, for now let's only use one site "A"
sitesToy <- rep(c("A"),length(timesToy))

#Then prepare some toy data:
#Prepare a noise and trend term to make the data more interesting
noise <- 0.05
trend <- 0.05

#Make cyclical data (note that we apply the trend only to x:
x <- sin((timesToy*2*pi)/cycleDurationToy)+rnorm(length(timesToy),mean=0,sd=noise)+trend*timesToy
y <- cos((timesToy*2*pi)/cycleDurationToy)+rnorm(length(timesToy),mean=0,sd=noise)
matToy <- cbind(x,y)

#And express it as a distance matrix (ETA is based on distances, increasing its generality)
dToy <- dist(matToy)
```

As with classical ETA, we define the cyclical trajectory using `defineTrajectories()`:
```{r}
xToy <- defineTrajectories(dToy, sites = sitesToy, times = timesToy)
```

Note that we did not define surveys, but added survey times as defined in `timesToy`. We can visualize the cyclical trajectory using the function `trajectoryPCoA()`, as we would do for any trajectory within ETA.
```{r fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
trajectoryPCoA(xToy, 
               lwd = 2,length = 0.2)
```

We see that the cyclical trajectory is composed of three cycles that are not yet explicitly identified. Similarly, the fixed-date trajectories are not isolated. This is the goal of the "extract" functions in CETA.

### 2.2 Extracting cycles and fixed-date trajectories

In CETA the "extract" functions take one or more **cyclical trajectories** and return a distance matrix `d` and associated descriptors allowing to isolate **cycles** or **fixed-date trajectories** (respectively functions `extractCycles()` and `extractFixedDateTrajectories()`) for further analysis.  

#### 2.2.1 Fixed-date trajectories

Let's start with `extractFixedDateTrajectories()` as it is the simplest one to handle:
```{r}
fdtrajToy <- extractFixedDateTrajectories(xToy,
                                          cycleDuration = cycleDurationToy,
                                          namesFixedDate = paste("M",1:10))
```
Note that we haven't specified  `datesToy` as a value for argument `dates`. This is because the CETA functions assume default values `times%%cycleDuration` which means the taking the modulo of cycle duration and works perfectly fine here. In cases there is an offset between dates and times (e.g. times = 0 does not correspond to dates = 0), a different `dates` argument must be provided.

Let's look at the output given by `extractFixedDateTrajectories()`:
```{r}
class(fdtrajToy)
```
The class `fd.trajectories` identifies the object as a subclass of `trajectories`. Its elements are the usual ones:
```{r}
names(fdtrajToy)
```
`d` is a distance matrix and `metadata` is a data frame containing the information necessary to read it as a descriptor of **fixed-date trajectories**.
```{r}
head(fdtrajToy$metadata)
```
The column `fdT` (for fixed-date trajectories) indicates to which fixed-date trajectories the different ecological states in `d` belong. The names in `fdT` are built by concatenating the site (especially useful if several cyclical trajectory are studied in parallel) with the string `fdT` and the name given as argument in `extractFixedDateTrajectories()` (if not provided, the date is used as default).

Using a combination of the new distance matrix `d` and its descriptors in `metadata`, the output of `extractFixedDateTrajectories()` can be fed in other ETA functions to study fixed-date trajectories. Whenever the ETA functions identify the input object as one of class `fd.trajectories`, they use the values of `fdT` column as substitute for `sites`. For instance we can compute the directionality of the fixed-date trajectories using:
```{r}
trajectoryDirectionality(fdtrajToy)
```

Or the distances between fixed-date trajectories using:
```{r}
trajectoryDistances(fdtrajToy)
```

For visualization, it is possible to use the general `trajectoryPCoA()` function as:
```{r fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
trajectoryPCoA(fdtrajToy,
               lwd = 2,length = 0.2,
               traj.colors = c("blue","red","black","grey","orange","green3","brown","purple","pink","yellow"))
```

But CETA provides a plotting function dedicated to fixed date trajectories that uses a circular color palette for the fixed dates trajectories and can represent the original cyclical trajectory:
```{r fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
fixedDateTrajectoryPCoA(fdtrajToy,
                        lwd = 2, length = 0.2)
```

Note that we find, as expected 10 fixed-date trajectories (one for each date of each cycle). They are all linear and pretty much parallel, which is what we expect given the trend that we put in our toy dataset.



#### 2.2.2 Trajectory cycles

Let's now look at how extraction works for **cycles**. The function `extractCycles()` works very similarly to `extractFixedDateTrajectories()`. Few differences are that: it does not allows to specify names for cycles (they are by default the concatenation of site names and "C1", "C2" etc...); it allows to change the `startdate` of the cycles (for instance allowing to flexibly choose when a cycle start: *e.g. an annual cycle does not always need to start in January*).
```{r}
cycleToy <- extractCycles(xToy,
                          cycleDuration = cycleDurationToy)
```

Just like `extractFixedDateTrajectories()`, `extractCycles()` has outputs `d` and `metadata`:
```{r}
names(cycleToy)
```

When inspecting `metadata`, we find the column `cycles`, containing the unique names of all cycles, analogously to column `fdT` in the output of `extractFixedDateTrajectories()`. In addition and importantly, `extractCycles()` returns the column named `internal`, which has no analog in `extractFixedDateTrajectories()`:
```{r}
head(cycleToy$metadata)
```

The `internal` column gives an information on *internal vs external* ecological states. This is an important distinction that solve what we call the **"December-to-January segment problem"** and that ‒ *unfortunately* ‒ complicates a bit the use of the outputs of `extractCycles()`. We explain this in the following section.

#### 2.2.3 Beware of the external ecological states: the "December-to-January segment problem"
Let's imagine we sampled a site monthly (Jan, Feb, ..., Dec) during many years ($Y$, $Y+1$ ...). How do we cut this cyclical trajectory into cycles? One possibility is to make cycles out of the segments joining all the months of year $Y$, from January to December (12 ecological states, joined by 11 segments). The problem then is that the segment joining December of year $Y$ to January of year $Y+1$ is ignored. We can then extend the cycles to the first ecological state of the next cycle (January of year $Y+1$, 13 ecological states 12 segments). This correctly includes the 12 segments in the cycle but implies that the cycle contains twice the month of January. We refer to this as the **"December-to-January segment problem"**.  

In CETA, we solve this issue by distinguishing "internal" and "external" ecological states. In the case above, January of year $Y+1$ would be considered "external" whereas other ecological states would be "internal". Broadly speaking, external ecological states are included in computations relying on the **cycle segments**, but excluded in computations relying on **cycle states**. More specifically the metrics and operations in CETA that require to remove, or apply a special treatment to external ecological states are:

* Centering, where internal states are used for determining cycle center but the centering operation applies to external states as well.
* PCoA for visualization.
* Computation of cycle smoothness or cyclical shifts, where external segments are required.
* Computation of trajectory variability.

In addition, you might have ideas to study cycles outside of the CETA/ETA framework. If so, please go ahead! But remember that most approaches rely on the concept of points, not segments, so external ecological states should probably be removed beforehand.  

We will detail how to treat external ecological states within CETA in the next sections, but first, let's explore the outputs of `extractCycles()` to make all this more palatable:
```{r}
nrow(as.matrix(dToy))#Number of ecological states in the original distance matrix.
nrow(as.matrix(cycleToy$d))#Number of ecological states in the matrix returned by extractCycles.
```
Note that, in this example, the distance matrix returned by `extractCycles()` describes two more ecological states than the original distance matrix. This is because `extractCycles()` duplicates ecological states if they are shared by two cycles (coming back to the example above, January of year $Y+1$ belongs to the cycle describing year $Y$ AND the cycle describing year $Y+1$). Such duplicated ecological states exist always in an "internal" and an "external" version.
This duplication allows to easily, and correctly, compute most of the ETA metrics for cycles such as, for instance, distances. Whenever they identify the input object as one of class `cycles`, they use the values of `cycles` column as substitute for `sites`. Akin to what we did with fixed-date trajectories, the key is to use the column `cycles` as input for `sites` in the ETA function :
```{r}
trajectoryDistances(cycleToy)
```
But if we want to center our cycles before computing the distances (for instance, if we are not interested in the trend but want to focus on changes of shape of the cycles) we need to take the distinction between internal and external ecological states in account and center all the distance matrix with respect to a centroid defined using ONLY internal ecological states. To do so we use the `internal` column. This column is a boolean set to `TRUE` if the associated ecological state is "internal":
```{r}
cycleToy_cent <- centerTrajectories(cycleToy,
                                    exclude = which(cycleToy$metadata$internal==FALSE))
```

In the previous call, the centering function was explicitly asked to exclude external ecological states from the centroid computation. After the centering the location of duplicated sites is no longer the same. The distances between cycles after centering are:

```{r}
trajectoryDistances(cycleToy_cent)
```

BLABLAH PCoA Cycles 

A special function is thus available within CETA for this purpose.
This can be done using the original cycles:
```{r fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
cyclePCoA(x = cycleToy, sites.colors = "blue")
```

As well as the centered cycles:
```{r fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
cyclePCoA(x = cycleToy_cent, centered = TRUE, sites.colors = "blue")
```

### 2.3 Assessing the degree of convexity in cycles
Another metric that need special handling in CETA is cycle smoothness, which is estimated using function `cycleConvexity()`. This metric is the cyclical equivalent of directionality for non-cyclical trajectories. A cycle with high convexity is interpreted as simpler (more convex) than a cycle with low convexity, which is due to concavities or "bends" in more than two dimension indicative of a more complex (or nosier) cyclical dynamic. The function `cycleConvexity()` does not uses the outputs of `extractCycles()`. This is because the original full cyclical trajectory is needed for computation. Instead, `cycleConvexity()` take the same inputs as `extractCycles()`:
```{r}
cycleConvexity(xToy,
               cycleDuration = cycleDurationToy)
```
Note that the function returns a `NA` for the first cycle. This is because the function needs to associate an angle to each internal ecological states of the cycles. Since the first cycle starts with the first ecological state of the whole time series, its associated angle cannot be computed, and neither the smoothness of the cycle.

### 2.4 Computing cyclical shifts: phenological advances and delays

A last novelty of CETA is to provide a way to generalize the concept of phenological advances and delays for multidimensional data such as those used in community ecology. The function that computes cyclical shifts is `cycleShifts()`. By default, it will estimate all the shifts that are possible to compute on a given cyclical trajectory. The computation relies on geometrical projection and comparison of a ecological states of interest to a reference cycle. The computation time can be long for large datasets. The function takes the same inputs as `extractCycles()`:
```{r}
cycleShifts(xToy, cycleDuration = cycleDurationToy)
```
The output gives the `cycleShifts()` computed in the same units as the input `time`, with positive values indicating advance and negative values indicating delay. The other columns give information on what exactly was compared. We will come back to this output in the real example below. 

### 2.5 Summary of the CETA approach
Here is a visual summary of the CETA approach:

![The CETA functions (blue) allow to format data to describe fixed-date trajectories or cycles. The reformated data (Output of functions `extractFixedDateTrajectories()` and `extractCycles()`) can then be used as inputs for other ETA functions (orange) to obtain the metrics of interest. However, when woking on cycles, keep in mind that external ecological states must be treated with care for some applications and that some metrics (obtained with functions `cycleConvexity` and `cycleShifts()`) are obtained directly from the original data, not from the output of `extractCycles()`.](CETA_workflow.png){width="100%"}

## 3. Real data example: Zooplankton of the North Sea 1958-2021

### 3.1 About the data
The data present here describes the zooplankton community in the North Sea sampled by the [Continuous Plankton Recorder (CPR)](https://www.cprsurvey.org/) survey. We re-worked the raw data provided by the CPR survey [(DOI: 10.17031/66f12be296d70)](https://doi.org/10.17031/66f12be296d70) into two monthly-resolved time series of the commonest zooplankton taxa in the Northern North Sea (`NNS`) and the Southern North Sea (`SNS`). In our data processing, we performed a smoothing by taking a rolling average (for each month, we averaged 5 values: a 3 months window + the corresponding month of the previous and next years). We finally Hellinger-transformed the abundance data to make them amenable to ecological diversity study. Hellinger transformation implies that only relative variations in abundances are studied here.

Let's call the data:
```{r load northseaZoo, echo=T}
data("northseaZoo")
```

```{r}
names(northseaZoo)
head(northseaZoo$times)
```
`northseaZoo` contains three elements: `Hellinger` with the Hellinger-transformed abundances, `times` with time expressed in years (the decimals actually correspond to a way of encoding the months as fractions of years: `1:12/12-1/24`), and `sites` describing whether a community sample is from the Northern `NNS` or Southern `SNS` North Sea.

### 3.2 Get a dissimilarity matrix and define cyclical trajectories

ETA works from dissimilarity matrices. So a first step in analyzing the North Sea zooplankton data is to go from the `Hellinger` matrix to a distance matrix. Luckily for us, euclidean distances computed on Hellinger-transformed community data yield Hellinger distances between communities. We can therefore use the `dist` function from base R.
```{r}
northseaZoo$Dist <- dist(northseaZoo$Hellinger)
```

To be analyzed using ETA, the distance matrix has to be complemented with information regarding sites and surveys to get a object of class `trajectories` using `defineTrajectories()`. 
```{r}
x_northseaZoo <- defineTrajectories(d = northseaZoo$Dist,
                                    sites = northseaZoo$sites,
                                    times = northseaZoo$times)
```


### 3.3 Visualize the cyclical trajectories

From now on, if you are executing the code live, bear with us a little bit as the time series are rather long, so your computer may need some time to display the graphics or compute some metrics.  
As before, we can use the `trajectoryPCoA()` function to display the two cyclical trajectories
```{r fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
trajectoryPCoA(x_northseaZoo,
               traj.colors = c("blue","red"),
               length = 0.05)
legend(x="topleft",col=c("blue","red"),pch=15,unique(northseaZoo$sites))
```
Things indeed seem to turn and present clear cycles: This is not surprising as seasonality is a very prominent factor in driving zooplankton community dynamics. But from such a representation it is hard to distinguish clear patterns. Let's use CETA!

### 3.4 Look at the cycles:

We can use the function `extractCycles()` to get data in a format describing the cycles:
```{r}
cyclesNSZoo <- extractCycles(x_northseaZoo,
                             cycleDuration = 1,
                             minEcolStates = 12)
#Note that we use here a cycleDuration of 1 (everything is expressed in years),
#we also use the minEcolStates argument to say that we want to only keep cycles that
#have at least 12 ecological states: In our case this is as complete as it gets
```

From there, we can use the outputs of `extractCycles()` contained in`cycleNSZoo` as inputs for other ETA function and assess some interesting characteristics of the seasonal cycles of North Sea zooplankton. For instance one can obtain cycle length:
```{r}
cyclesZooLengths <- trajectoryLengths(cyclesNSZoo)
```

However, remember that cycle convexity is not obtained using the output of `extractCycles()`, instead we get it using the same inputs as for `extractCycles()` such as:
```{r}
cyclesZooSmooth <- cycleConvexity(x_northseaZoo,
                                  cycleDuration = 1,
                                  minEcolStates = 12)
```

Cycle length and cycle smoothness help describing the shapes of cycles in a way mirroring trajectory length and directionality in non-cyclical ETA. Let's graph the outputs:
```{r}
#First let's build some broad descriptive statistics for individual cycles: what year, and what site? (MAYBE THAT COULD BE DONE AUTOMATICALLY BY extractCycles???):
yearCycles <- floor(tapply(cyclesNSZoo$metadata$times,cyclesNSZoo$metadata$cycles,min))
sitesCycles <- tapply(cyclesNSZoo$metadata$sites,cyclesNSZoo$metadata$cycles,unique)

#Doing some reordering
yearCycles <- yearCycles[unique(cyclesNSZoo$metadata$cycles)]
sitesCycles <- sitesCycles[unique(cyclesNSZoo$metadata$cycles)]

#put all descriptors together
StatCyclesZoo <- data.frame(sitesCycles,yearCycles,cyclesZooLengths$Trajectory,cyclesZooSmooth)
SNScycles <- subset(StatCyclesZoo,sitesCycles=="SNS")
NNScycles <- subset(StatCyclesZoo,sitesCycles=="NNS")

#put that in the format of a complete time serie (recreating the holes if any)
rownames(SNScycles) <- SNScycles$yearCycles
rownames(NNScycles) <- NNScycles$yearCycles

SNScycles <- SNScycles[as.character(1958:2021),]
NNScycles <- NNScycles[as.character(1958:2021),]
```

```{r fig = TRUE, fig.height=5, fig.width=7, fig.align = "center"}
#First figure: Lengths
plot(SNScycles$yearCycles,SNScycles$cyclesZooLengths.Trajectory,type="l",las=1,ylab="Cycle lengths",xlab="Years",ylim=c(2.5,4.1),col="red")
points(NNScycles$yearCycles,NNScycles$cyclesZooLengths.Trajectory,type="l",col="blue")
```

```{r fig = TRUE, fig.height=5, fig.width=7, fig.align = "center"}
#Second figure: Smoothness
plot(SNScycles$yearCycles,SNScycles$cyclesZooSmooth,type="l",las=1,ylab="Cycle smoothness",xlab="Years",ylim=c(0.3,0.5),col="red")
points(NNScycles$yearCycles,NNScycles$cyclesZooSmooth,type="l",col="blue")
```

Cycle length may not tell much with those data, but in smoothness, there seems to be a "jump" around year 2000. From previous studies on the CPR time series, it is known that echinoderm larvae became seasonally dominant around this time in the North Sea. It is possible that this "jump" is linked to this event, a great dominance of one taxa in a given season would indeed pull the cycles towards one dimension only, possibly reducing concavities and thus increasing smoothness.

Let's now look at the distances between cycles:
```{r}
cyclesZooDistances <- trajectoryDistances(cyclesNSZoo)
```

Distances between cycles can then be visualized by principal coordinates analysis (PCoA) putting us in the "space of cycles":
```{r}
library(ape)
PCoAZoo <- pcoa(cyclesZooDistances)
```

```{r fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
colPoints <- c("red","blue")
names(colPoints) <- c("SNS","NNS")
plot(PCoAZoo$vectors[,1:2],asp=1,col=NA,
     xlab=paste("PCoA 1 (",round(PCoAZoo$values$Relative_eig[1]*100,2)," %)",sep=""),
     ylab=paste("PCoA 2 (",round(PCoAZoo$values$Relative_eig[2]*100,2)," %)",sep=""))
text(PCoAZoo$vectors[,1:2],as.character(yearCycles),col=colPoints[sitesCycles])
legend(x="topleft",col=c("blue","red"),pch=15,unique(northseaZoo$sites))
```

We can see from this graph that the Northern and Southern North Sea present parallel changes in seasonal dynamics while maintaining their differences during the whole time series. An important change seem to have occurred around year 2000 in agreement with what has already been said from those data using other methods.  

You know what's fun? We now have two sets of time-ordered "ecological states" (more ecological dynamics to be fair) defined in a distance matrix. This is all we need to do some trajectory analysis. We can thus perform what we call "second-order" trajectory analysis using the distance between cycles as our distance matrix. We wont go in the details here but here is the corresponding graph as an appetizer:
```{r fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
x_second_order <- defineTrajectories(d = cyclesZooDistances,
                                     sites = sitesCycles,
                                     surveys = yearCycles)
trajectoryPCoA(x_second_order,
               traj.colors=c("blue","red"),
               length=0.05)
legend(x="topleft",col=c("blue","red"),pch=15,unique(northseaZoo$sites))
```

Note that the two PCoA do not represent the same variance on their first and second axes. This is due to differences in the correction method used for negative eigenvalues (ignored in the first case, application of Cailliez correction in the second case).

### 3.5 Look at the fixed-date trajectories:

Let's now look at the fixed-date trajectories. First prepare the data using `extractFixedDateTrajectories()`:
```{r}
fdtrajNSZoo <- extractFixedDateTrajectories(x_northseaZoo,
                                            cycleDuration = 1,
                                            namesFixedDate =
                             c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec"))
```

As we did for cycles, we can compute shape metrics on fixed-date trajectories:
```{r}
fdtrajZooLengths <- trajectoryLengths(fdtrajNSZoo)
fdtrajZooDir <- trajectoryDirectionality(fdtrajNSZoo)
```

And put them in a convenient dataframe:
```{r}
#descriptive stats (MAYBE THAT COULD BE DONE AUTOMATICALLY BY extractFixedDateTrajectories???):
monthFDT <- tapply(fdtrajNSZoo$metadata$dates,fdtrajNSZoo$metadata$fdT,min)
sitesFDT <- tapply(fdtrajNSZoo$metadata$sites,fdtrajNSZoo$metadata$fdT,unique)

#reordering
monthFDT <- monthFDT[unique(fdtrajNSZoo$metadata$fdT)]
sitesFDT <- sitesFDT[unique(fdtrajNSZoo$metadata$fdT)]

#put all descriptors together
StatFDTZoo <- data.frame(sitesFDT,monthFDT,fdtrajZooLengths$Trajectory,fdtrajZooDir)
SNSfdT <- subset(StatFDTZoo,sitesFDT=="SNS")
NNSfdT <- subset(StatFDTZoo,sitesFDT=="NNS")

#put them in month order
SNSfdT <- SNSfdT[order(SNSfdT$monthFDT),]
NNSfdT <- NNSfdT[order(NNSfdT$monthFDT),]

#and put the month in a convenient format:
SNSfdT$monthFDT <- SNSfdT$monthFDT*12+0.5
NNSfdT$monthFDT <- NNSfdT$monthFDT*12+0.5
```

Before graphing them:
```{r fig = TRUE, fig.height=5, fig.width=7, fig.align = "center"}
#First figure: Lengths
plot(SNSfdT$monthFDT,SNSfdT$fdtrajZooLengths.Trajectory,type="l",las=1,ylab="Fixed-date trajectory lengths",xlab="Month",ylim=c(8,22),col="red")
points(NNSfdT$monthFDT,NNSfdT$fdtrajZooLengths.Trajectory,type="l",col="blue")
```

```{r fig = TRUE, fig.height=5, fig.width=7, fig.align = "center"}
#Second figure: Directionality
plot(SNSfdT$monthFDT,SNSfdT$fdtrajZooDir,type="l",las=1,ylab="Fixed-date trajectory directionality",xlab="Years",ylim=c(0.36,0.47),col="red")
points(NNSfdT$monthFDT,NNSfdT$fdtrajZooDir,type="l",col="blue")
```

There is tendency of fixed-date trajectories to have high directionality and low length in summer months and conversely in winter. This is likely to be a sampling effect. Although the sampling effort is rather continuous throughout the year in CPR data, winter communities are less abundant making their measured composition more stochastic. This is reflected in more noisy fixed-date trajectories in winter.

Nonetheless, a metric of peculiar interest in fixed-date trajectories might be convergence:
```{r}
#We will do it for the two sites (NNS and SNS) separately, so lets pull them apart
SNSfdtraj <- fdtrajNSZoo
NNSfdtraj <- fdtrajNSZoo

#change the distance matrices
SNSfdtraj$d <- as.dist(as.matrix(SNSfdtraj$d)[SNSfdtraj$metadata$sites=="SNS",SNSfdtraj$metadata$sites=="SNS"])
NNSfdtraj$d <- as.dist(as.matrix(NNSfdtraj$d)[NNSfdtraj$metadata$sites=="NNS",NNSfdtraj$metadata$sites=="NNS"])

#change metadata
SNSfdtraj$metadata <- SNSfdtraj$metadata[SNSfdtraj$metadata$sites=="SNS",]
NNSfdtraj$metadata <- NNSfdtraj$metadata[NNSfdtraj$metadata$sites=="NNS",]

#Then we need to keep only the years during which the fixed-date trajectories all have associated ecological states (this is because we want to perform a symmetric convergence test)
selecSNS <- as.numeric(names(which(table(SNSfdtraj$metadata$times-SNSfdtraj$metadata$dates)==12)))
selecNNS <- as.numeric(names(which(table(NNSfdtraj$metadata$times-NNSfdtraj$metadata$dates)==12)))

#change the distance matrices again
SNSfdtraj$d <- as.dist(as.matrix(SNSfdtraj$d)[floor(SNSfdtraj$metadata$times)%in%selecSNS,floor(SNSfdtraj$metadata$times)%in%selecSNS])
NNSfdtraj$d <- as.dist(as.matrix(NNSfdtraj$d)[floor(NNSfdtraj$metadata$times)%in%selecNNS,floor(NNSfdtraj$metadata$times)%in%selecNNS])

# and change metadata again
SNSfdtraj$metadata <- SNSfdtraj$metadata[floor(SNSfdtraj$metadata$times)%in%selecSNS,]
NNSfdtraj$metadata <- NNSfdtraj$metadata[floor(NNSfdtraj$metadata$times)%in%selecNNS,]

#Now we can compute the convergence!
SNSfdtrajConv <- trajectoryConvergence(SNSfdtraj,
                                       symmetric = TRUE)
NNSfdtrajConv <- trajectoryConvergence(NNSfdtraj,
                                       symmetric = TRUE)

library(corrplot)
```

Let's now visualize it for SNS:
```{r fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
corrplot(matrix(as.vector(SNSfdtrajConv$tau)*as.numeric(SNSfdtrajConv$p.value<0.05),12,12))
```
And for NNS:
```{r fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
corrplot(matrix(as.vector(NNSfdtrajConv$tau)*as.numeric(NNSfdtrajConv$p.value<0.05),12,12))
```
In those graphs, blue indicates divergence while red indicates convergence. In both part of the North Sea, we see a convergence of spring and summer months and a divergence winter month with summer months. One possible interpretation is that summers and winters are getting more contrasted.

### 3.6 Finally, compute cyclical shifts:

The last aspect that CETA allows to investigate is cyclical shifts (e.g. advances and delays similar to approaches in phenology). The function that does it is `cycleShifts()`. This one will take a bit long to compute, it works a lot!
```{r}
CSNSZoo <- cycleShifts(x_northseaZoo,
                       cycleDuration = 1)
```

We can then extract a trend for each month in each site (and in the process make one example graph):
```{r fig = TRUE, fig.height=5, fig.width=7, fig.align = "center"}
slopes <- integer(0)
for (i in c("SNS","NNS")){
  for (j in unique(CSNSZoo$dateCS)){
    #prepare the relevant subset
    subsetCS <- subset(CSNSZoo,sites==i)|>subset(dateCS==j)
    
    #compute a slope
    model <- lm((subsetCS$cyclicalShift*365)~subsetCS$timeScale)
    #Note that we multiply the shift by 365 to get in days
    sum <- summary(model)
    slopes <- rbind(slopes,data.frame(i,j,model$coefficients[2]))
    
    if (i=="NNS"&j==unique(CSNSZoo$dateCS)[5]){
      plot(x=subsetCS$timeScale,y=subsetCS$cyclicalShift*365,ylim=c(-182.5,182.5),
       col=rgb(0,0,0,0.2),
       las=1,
       pch=16,
       xlab="Time scale (Years)",
       ylab="Seasonal offset (days)",
       main="Cyclical shifts for May in NNS")
      abline(h=0)
      abline(model,lwd=2,col="orangered")
      moys <- tapply(subsetCS$cyclicalShift*365,subsetCS$timeScale,mean)
      points(x=as.numeric(names(moys)),y=moys,pch=21,type="b",bg="dodgerblue2",cex=1.2)
    }
  }
}
colnames(slopes) <- c("sites","month","slope")

```

The graph gives an example of how trends of cyclical shifts can be computed across the whole time series for a given month. Let's now synthesize this information first for SNS:
```{r fig = TRUE, fig.height=5, fig.width=7, fig.align = "center"}
plot(x=slopes$month[slopes$sites=="SNS"]*12+0.5,
     y=slopes$slope[slopes$sites=="SNS"]*10,
     las=1,type="b",
     ylab="Cyclical shift in SNS (days/decade)",
     xlab="Month")
#note the multiplication by ten to get a slope in days per decade
abline(h=0)

```

Same for the NNS
```{r fig = TRUE, fig.height=5, fig.width=7, fig.align = "center"}
plot(x=slopes$month[slopes$sites=="NNS"]*12+0.5,
     y=slopes$slope[slopes$sites=="NNS"]*10,
     las=1,type="b",
     ylab="Cyclical shift in NNS (days/decade)",
     xlab="Month")
#note the multiplication by ten to get a slope in days per decade
abline(h=0)

```

The two region of the North Sea exhibit a similar pattern of cyclical shift: there is maximal advance (positive shift) in spring and autumn. The advance computed is not negligible, communities take up to 6-10 days of advance every decade at those months (at least in term of composition)!

