---
title: "Distance metrics for trajectory resemblance"
author: "Miquel De Cáceres"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: TRUE
vignette: >
  %\VignetteIndexEntry{Distance metrics for trajectory resemblance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignettePackage{ecotraj}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction

In this vignette you will learn the differences between the distance coefficients available for assessing trajectory resemblance. We use small datasets where trajectories occur in a space of two dimensions, so that geometric calculations can be followed more easily. 
First of all, we load `ecotraj`:

```{r load libraries, echo = T}
library(ecotraj)
```


## 2. Linear trajectories

### 2.1 Oposed linear trajectories
```{r}
sites <- c("1","1","1","1","2","2","2", "2")
times <- c(0,1,2,3,0,1,2,3)
  
xy<-matrix(0, nrow=8, ncol=2)
xy[2,2]<-1
xy[3,2]<-2
xy[4,2]<-3
xy[5,2]<-3
xy[6,2]<-2
xy[7,2]<-1
xy[8,2]<-0

x <- defineTrajectories(dist(xy), sites = sites, times = times)
```


```{r, fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
trajectoryPCoA(x, 
               traj.colors = c("black", "red", "blue"), lwd = 2,
               time.labels = TRUE)
```
```{r}
trajectoryLengths(x)
```

```{r}
trajectorySpeeds(x)
```


```{r}
trajectoryShifts(x)
```

```{r}
trajectoryDistances(x, distance.type = "SPD")
```

```{r}
trajectoryDistances(x, distance.type = "DSPD")
```


```{r}
trajectoryDistances(x, distance.type = "TSD")
```

### 2.2 Equal trajectories but different number of segments
```{r}
sites <- c("1","1","1","1","2","2","2","3","3")
times <- c(0,1,2,3,0,1.5,3,0,3)
  
xy<-matrix(0, nrow=9, ncol=2)
xy[2,2]<-1
xy[3,2]<-2
xy[4,2]<-3
xy[6,2]<-1.5
xy[7,2]<-3
xy[9,2]<-3

x <- defineTrajectories(dist(xy), sites = sites, times = times)
```

```{r, fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
trajectoryPCoA(x, 
               traj.colors = c("black", "red", "blue"), lwd = 2,
               time.labels = TRUE)
```
```{r}
trajectoryLengths(x)
```

```{r}
trajectorySpeeds(x)
```


```{r}
trajectoryShifts(x)
```

```{r}
trajectoryDistances(x, distance.type = "SPD")
```

```{r}
trajectoryDistances(x, distance.type = "DSPD")
```

```{r}
trajectoryDistances(x, distance.type = "TSD")
```


### 2.3 Equal pathways but different speeds


```{r}
sites <- c("1","1","1","1","2","2","2","2","3","3","3","3")
times <- c(0,0.5,1,1.5,0,1,2,3,0,2,4,6)
  
xy<-matrix(0, nrow=12, ncol=2)
xy[2,2]<-1
xy[3,2]<-2
xy[4,2]<-3
xy[5:8,2] <- xy[1:4,2]
xy[9:12,2] <- xy[1:4,2]

x <- defineTrajectories(dist(xy), sites = sites, times = times)
```

```{r, fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
par(mfrow=c(3,1))
trajectoryPCoA(subsetTrajectories(x,"1"), 
               traj.colors = c("black"), lwd = 2,
               time.labels = TRUE)
trajectoryPCoA(subsetTrajectories(x,"2"), 
               traj.colors = c("red"), lwd = 2,
               time.labels = TRUE)
trajectoryPCoA(subsetTrajectories(x,"3"), 
               traj.colors = c("blue"), lwd = 2,
               time.labels = TRUE)
```

```{r}
trajectoryLengths(x)
```

```{r}
trajectorySpeeds(x)
```

```{r}
trajectoryShifts(x)
```



```{r}
trajectoryDistances(x, distance.type = "SPD")
```

```{r}
trajectoryDistances(x, distance.type = "DSPD")
```

```{r}
trajectoryDistances(x, distance.type = "TSD")
```

### 2.4 Space-shifted trajectories

```{r}
sites <- c("1","1","1","1","2","2","2","2","3","3","3","3")
times <- c(1,2,3,4,1,2,3,4,1,2,3,4)

xy<-matrix(0, nrow=12, ncol=2)
xy[2,2]<-1
xy[3,2]<-2
xy[4,2]<-3
# States are all shifted half unit with respect to site "1"
xy[5:8,2] <- xy[1:4,2] + 0.5 
# States are all shifted one unit with respect to site "1"
xy[9:12,2] <- xy[1:4,2] + 1.0  

x <- defineTrajectories(dist(xy), sites = sites, times = times)
```



```{r, fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
trajectoryPCoA(x, 
               traj.colors = c("black", "red", "blue"), lwd = 2,
               time.labels = TRUE)
```
```{r}
trajectoryLengths(x)
```

```{r}
trajectorySpeeds(x)
```


```{r}
trajectoryShifts(x)
```

```{r}
trajectoryDistances(x, distance.type = "SPD")
```

```{r}
trajectoryDistances(x, distance.type = "DSPD")
```

```{r}
trajectoryDistances(x, distance.type = "TSD")
```

### 2.5 Space-expanded trajectories

```{r}
sites <- c("1","1","1","1","2","2","2","2","3","3","3","3")
times <- c(0,1,2,3,0,1,2,3,0,1,2,3)
  
xy<-matrix(0, nrow=12, ncol=2)
xy[2,2]<-1
xy[3,2]<-2
xy[4,2]<-3
xy[5:8,2] <- xy[1:4,2]*1.5
xy[9:12,2] <- xy[1:4,2]*2

x <- defineTrajectories(dist(xy), sites = sites, times = times)
```

```{r, fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
trajectoryPCoA(x, 
               traj.colors = c("black", "red", "blue"), lwd = 2,
               time.labels = TRUE)
```

```{r}
trajectoryLengths(x)
```

```{r}
trajectorySpeeds(x)
```


```{r}
trajectoryShifts(x)
```

```{r}
trajectoryDistances(x, distance.type = "SPD")
```

```{r}
trajectoryDistances(x, distance.type = "DSPD")
```

```{r}
trajectoryDistances(x, distance.type = "TSD")
```

## 3. Curved trajectories

### 3.1 Constant speed

```{r}
sites <- c("1","1","1","1","2","2","2","2","3","3","3","3")
surveys <- c(1,2,3,4,1,2,3,4,1,2,3,4)

xy<-matrix(0, nrow=12, ncol=2)
xy[2,2]<-1
xy[3,2]<-2
xy[4,2]<-3
xy[5:6,2] <- xy[1:2,2]
xy[7,1]<-0+sqrt(0.5)
xy[7,2]<-1+sqrt(0.5)
xy[8,2]<-xy[7,2]
xy[8,1]<-xy[7,1]+1
xy[9:10,2] <- xy[1:2,2]
xy[11,1] <- 1.0
xy[11,2] <- 1.0
xy[12,1] <- 1.0
xy[12,2] <- 0.0

x <- defineTrajectories(dist(xy), sites = sites, times = times)
```


```{r, fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
trajectoryPCoA(x, 
               traj.colors = c("black", "red", "blue"), lwd = 2,
               time.labels = TRUE)
```

```{r}
trajectoryLengths(x)
```

```{r}
trajectorySpeeds(x)
```


```{r}
trajectoryShifts(x)
```

```{r}
trajectoryDistances(x, distance.type = "SPD")
```

```{r}
trajectoryDistances(x, distance.type = "DSPD")
```

```{r}
trajectoryDistances(x, distance.type = "TSD")
```

### 3.2 Different speed


```{r}
sites <- c("1","1","1","1","2","2","2","2","3","3","3","3")
times <- c(0,0.5,1,1.5,0,1,2,3,0,2,4,6)

xy<-matrix(0, nrow=12, ncol=2)
xy[2,2]<-1
xy[3,2]<-2
xy[4,2]<-3
xy[5:6,2] <- xy[1:2,2]
xy[7,1]<-0+sqrt(0.5)
xy[7,2]<-1+sqrt(0.5)
xy[8,2]<-xy[7,2]
xy[8,1]<-xy[7,1]+1
xy[9:10,2] <- xy[1:2,2]
xy[11,1] <- 1.0
xy[11,2] <- 1.0
xy[12,1] <- 1.0
xy[12,2] <- 0.0

x <- defineTrajectories(dist(xy), sites = sites, times = times)
```


```{r, fig = TRUE, fig.height=5, fig.width=5, fig.align = "center"}
trajectoryPCoA(x, 
               traj.colors = c("black", "red", "blue"), lwd = 2,
               time.labels = TRUE)
```

```{r}
trajectoryLengths(x)
```

```{r}
trajectorySpeeds(x)
```


```{r}
trajectoryShifts(x)
```

```{r}
trajectoryDistances(x, distance.type = "SPD")
```

```{r}
trajectoryDistances(x, distance.type = "DSPD")
```

```{r}
trajectoryDistances(x, distance.type = "TSD")
```
